Equivalence Testing and the Second Generation \emph{P}-Value

### 2018-08-28 manuscript.tex
submission

@@

To test predictions researchers predominantly rely on null-hypothesis
tests. This statistical approach can be used to examine whether observed
data is sufficiently surprising under the null hypothesis to reject an
effect that equals exactly zero. Null-hypothesis tests have an important
limitation, in that this procedure can only reject the hypothesis that
there is no effect, while scientists should also be able to provide
statistical support for \emph{equivalence}. When testing for equivalence
researchers aim to examine whether an observed effect is too small to be
considered meaningful, and therefore is practically equivalent to zero.
By specifying a range around the null hypothesis of values that are
deemed practically equivalent to the absence of an effect (i.e., 0 ±
0.3) the observed data can be compared against an \emph{equivalence
range} and researchers can test if a meaningful effect is absent (Hauck
\& Anderson, 1984; Kruschke, 2018; Rogers, Howard, \& Vessey, 1993;
Serlin \& Lapsley, 1985; Spiegelhalter, Freedman, \& Parmar, 1994;
Wellek, 2010; Westlake, 1972).

@@

The TOST procedure also relies on the confidence interval around the
effect. In the TOST procedure the data is tested against the lower
equivalence bound in the first one-sided test, and against the upper
equivalence bound in the second one-sided test (Lakens, Scheel, \&
Isager, 2018). If both tests statistically reject an effect as extreme
or more extreme than the equivalence bound, you can conclude the
observed effect is practically equivalent to zero from a Neyman-Pearson
approach to statistical inferences. Because one-sided tests are
performed, one can also conclude equivalence by checking whether the
1-2\(\times\)alpha confidence interval (e.g., when the alpha level is
0.05, a 90\% CI) falls completely within the equivalence bounds. Because
both equivalence tests as the SGPV are based on whether and how much a
confidence interval overlaps with equivalence bounds, it seems
worthwhile to compare the behavior of the newly proposed SGPV to
equivalence tests to examine the unique contribution of the SGPV to the
statistical toolbox.

@@

Two other points always have to overlap. When the 95\% CI falls
completely inside the equivalence region, and one endpoint of the
confidence interval is exactly equal to one of the equivalence bounds
(see situation B in Figure \ref{fig:TOSTSGPV3}) the TOST \emph{p}-value
(which relies on a one-sided test) is always 0.025. The third point
where the SGPV and the \emph{p}-value from the TOST procedure should
overlap is where the 95\% CI falls completely outside of the equivalence
range, but one endpoint of the confidence interval is equal to the
equivalence bound (see situation C in Figure \ref{fig:TOSTSGPV3}), when
the \emph{p}-value will always be 0.975. Note that this situation is in
essence a \emph{minimum-effect test} (Murphy, Myors, \& Wolach, 2014).
The goal of a minimum-effect is not just to reject a difference of zero,
but to reject the smallest effect size of interest (i.e., the
equivalence bounds). The SGPV summarizes the information from an
equivalence test and a minimum-effect test. These can be two relevant
questions to ask, although it often makes sense to combine an
equivalence test and a null-hypothesis test instead (Lakens et al.,
2018).

@@

We can examine this situation by calculating the SGPV and performing the
TOST for a situation where sample sizes are small and the equivalence
range is narrow, such that the CI is more than twice as large as the
equivalence range (see Figure \ref{fig:TOSTSGPV6}). When the two
statistics are plotted against each other we can seewhere they are
unrelated (indicated by straight lines in the curve, see Figure
\ref{fig:TOSTSGPV7}). We see the SGPV is 0.5 for a range of observed
means where the \emph{p}-value from the equivalence test still varies.
It should be noted that in these calculations the \emph{p}-values for
the TOST procedure are \emph{never} smaller than 0.05 (i.e., they do not
get below 0.05 on the y-axis). In other words, we cannot conclude
equivalence based on any of the observed means. This happens because the
TOST \emph{p}-value is smaller than 0.05 only when the 90\% CI falls
completely between the upper and lower equivalence bounds. However, we
are examining a scenario where the 90\% CI is so wide that it never
falls completely within the two equivalence bounds, and thus the
equivalence test is never significant. As Lakens (2017) notes:
\enquote{in small samples (where CIs are wide), a study might have no
statistical power (i.e., the CI will always be so wide that it is
necessarily wider than the equivalence bounds).} None of the
\emph{p}-values based on the TOST procedure are below 0.05, and thus, in
the long run we have 0\% power.

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/TOSTSGPV9-1.pdf}
\caption{\label{fig:TOSTSGPV9}Comparison of \emph{p}-values from TOST (black
line) and SGPV (dotted grey line) across a range of observed sample
means (x-axis). The sample size is small (n = 10), but because the sd is
half as big as in Figure \ref{fig:TOSTSGPV7} (1 instead of 2) the CI is
less than twice as wide as the equivalence range (set to -0.4 to 0.4).
The SGPV is not set to 0.5 (horizontal lightgrey line) but reaches a
maximum slightly above 0.5 across a range of observed means.}
\end{figure}

@@

If we once more plot the two statistics against each other to see where
they are unrelated (indicated by straight lines in the curve), we see
the SGPV is 0.56 for a range of observed means where the \emph{p}-value
from the equivalence test still varies (Figure \ref{fig:TOSTSGPV10}). To
conclude this section, there are three situations where the
\emph{p}-value from the TOST procedure is unrelated to the SGPV. In all
these situations the \emph{p}-value for the equivalence test
differentiates tests with different means, but the SGPV does not.
Therefore, as a purely continuous descriptive statistic, the SGPV is
more limited than the \emph{p}-value from the TOST procedure.

@@

This example illustrates the difference between a proportion and a
probability. There is always a 50\% probability of observing a
correlation smaller or larger than the true correlation, but the SGPV
for this situation depends on how far away the observed correlation is
from 0. The further away from 0, the larger the SGPV when the observed
mean falls on the equivalence bound. The SGPV is the proportion of
values in a 95\% confidence interval that overlap with the equivalence
range, but not the probability that these values will be observed. In
the most extreme case (i.e., a sample size of 4, and equivalence bounds
set to r = -0.99 and 0.99, with an observed correlation of 0.99) 97.60\%
of the confidence interval overlaps with the equivalence range, even
though in the long run only 50\% of the correlations observed in the
future will fall in this range. It should be noted that in larger sample
sizes the SGPV is closer to 0.5 whenever the observed correlation falls
on the equivalence bound, but this extreme example nevertheless clearly
illustrates the difference between question the SGPV answers, and the
question a \emph{p}-value answers. The conclusion of this section on
asymmetric confidence intervals is that a SGPV of 1 or 0 can still be
interpreted as a \emph{p} \textless{} 0.025 or \emph{p} \textgreater{}
0.975 in an equivalence test, since the SGPV and \emph{p}-value for the
TOST procedure are always directly related at these values. Although
Blume et al. (2018) state that \enquote{the degree of overlap conveys
how compatible the data are with the null premise} this definition of
what the SGPV provides does not hold for asymmetric confidence
intervals. Although a SGPV of 1 or 0 can be directly interpreted, a SGPV
between 0 and 1 is not interpretable as \enquote{compatibility with the
null hypothesis}. Indeed, Blume and colleagues write in the supplemental
material that \enquote{The magnitude of an inconclusive
second-generation \emph{p}-value can vary slightly when the effect size
scale is transformed. However definitive findings, i.e.~a \emph{p}-value
of 0 or 1 are \emph{not} affected by the scale changes.}

@@

When introducing a new statistical method, it is important to compare it
to existing approaches and specify it's relative strengths and
weaknesses. First of all, the SGPV is a descriptive statistic (unlike
the \emph{p}-value that is calculated for an equivalence test, which is
an inferential statistic) that provides the proportion of overlap of the
confidence interval and the equivalence range. Even though a SGPV of 1
or 0 has a clear interpretation (we can reject effects outside or inside
the equivalence range), intermediate values are not as easy to interpret
(especially for effects that have asymmetric confidence intervals). In
one sense, they are what they are (the proportion of overlap), but it
can be unclear what this number tells us about the data we have
collected. This is not too problematic, since the main use of the SGPV
(e.g., in all examples provided by Blume and colleagues) seems to be to
examine whether the SGPV is 0, 1, or inconclusive. This interpretation
of a SGPV as allowing researchers to reject the null, reject the
presence of a meaningful effect, or remaining inconclusive is very
similar to the Neyman-Pearson interpretation of combining a
null-hypothesis test and an equivalence test (Lakens et al., 2018). The
difference is that where a SGPV of 1 can be interpreted as \emph{p}
\textless{} .025, equivalence tests provide exact \emph{p}-values, and
they continue to differentiate between for example \emph{p} = 0.048 and
\emph{p} = 0.002. Whether this is desirable depends on the perspective
that is used. From a Neyman-Pearson perspective on statistical
inferences the main conclusion is based on whether or not
\(p < \alpha\), and thus an equivalence test and SGPV can be performed
by simply checking whether the confidence interval falls within the
equivalence range, just as a null-hypothesis test can be performed by
checking whether the confidence interval contains zero or not. At the
same time, it is recommended to report exact \emph{p}-values (American
Psychological Association, 2010), and exact \emph{p}-values might
provide information of interest to readers about how precisely how
surprising the data is under the null model. Equivalence tests combined
with null-hypothesis significance tests also allow researchers to
conclude an effect is significant \emph{and} equivalent (i.e.,
statistically different from zero, but also too small to be considered
meaningful). Thus, the SGPV is used to classify results into one of
three possible categories (with the data falling inside or outside the
equivalence range, or being inconclusive), while equivalence tests
combined with null-hypothesis tests classify results into four possible
categories.

@@

As a more extreme example of the peculiar behavior of the \enquote{small
sample correction} as currently implemented in the calculation of the
SGPV see Figure \ref{fig:TOSTSGPV13} below. In this figure observed
correlations (from a sample size of 10) from -1 to 1 are tested against
an equivalence range from r = 0.4 to r = 0.8. We can see the SGPV has a
peculiar shape because it is set to 0.5 for certain observed
correlations, even though there is no risk of a misleading SGPV in this
range. This example suggests that the current implementation of the
\enquote{small sample correction} could be improved. If, on the other
hand, the SGPV is mainly meant to be interpreted when it is 0 or 1, it
might be preferable to simply never apply the \enquote{small sample
correction}.

Blume et al. (2018) argue that the SGPV has improved error control, in
that the conclusion that the 95\% confidence interval lies completely
outside the equivalence range around zero will necessarily occur less
frequently than a Type I error in a null-hypothesis test (where the 95\%
confidence interval only needs to not overlap with zero). However, the
SGPV has a \emph{lower} error rate than a null-hypothesis test, not a
\emph{more accurate} error rate. In a Neyman-Pearson perspective (which
forms the basis of equivalence tests) the goal is not to end up with an
error rate that is as low as possible (to achieve that, one simply
adjusts the alpha level), but with a decision procedure that, when
applied, yields a desired error rate with high accuracy. We show here
that the SGPV and equivalence tests are directly related, and so are
their error rates. The TOST procedure uses a 90\% confidence interval
because it is based on two \emph{one-sided} tests, and thus the
confidence interval is 1-2*\(\alpha\), and because both tests need to be
significant to conclude equivalence, no adjustment for multiple
comparisons is needed. The TOST procedure always has higher power to
declare equivalence than the SGPV, which relies on a 95\% confidence
interval (but note that there is no reason to always set the alpha level
to 0.05).

>>>>

 % Line numbering
  \usepackage{lineno}
  \linenumbers

@@

To test predictions researchers predominantly rely on null-hypothesis
tests.\\
This statistical approach can be used to examine whether observed data
is sufficiently surprising under the null hypothesis to reject an effect
that equals exactly zero. Null-hypothesis tests have an important
limitation, in that this procedure can only reject the hypothesis that
there is no effect, while scientists should also be able to provide
statistical support for \emph{equivalence}. When testing for equivalence
researchers aim to examine whether an observed effect is too small to be
considered meaningful, and therefore is practically equivalent to zero.
By specifying a range around the null hypothesis of values that are
deemed practically equivalent to the absence of an effect (i.e., 0 ±
0.3) the observed data can be compared against an \emph{equivalence
range} and researchers can test if a meaningful effect is absent (Hauck
\& Anderson, 1984; Kruschke, 2018; Rogers, Howard, \& Vessey, 1993;
Serlin \& Lapsley, 1985; Spiegelhalter, Freedman, \& Parmar, 1994;
Wellek, 2010; Westlake, 1972).

@@

The TOST procedure also relies on the confidence interval around the
effect. In the TOST procedure the data is tested against the lower
equivalence bound in the first one-sided test, and against the upper
equivalence bound in the second one-sided test (Lakens, Scheel, \&
Isager, 2018). If both tests statistically reject an effect as extreme
or more extreme than the equivalence bound, you can conclude the
observed effect is practically equivalent to zero from a Neyman-Pearson
approach to statistical inferences. Because one-sided tests are
performed, one can also conclude equivalence by checking whether the
1-2\(\times\)\(\alpha\) confidence interval (e.g., when the alpha level
is 0.05, a 90\% CI) falls completely within the equivalence bounds.
Because both equivalence tests as the SGPV are based on whether and how
much a confidence interval overlaps with equivalence bounds, it seems
worthwhile to compare the behavior of the newly proposed SGPV to
equivalence tests to examine the unique contribution of the SGPV to the
statistical toolbox.

@@

Two other points always have to overlap. When the 95\% CI falls
completely inside the equivalence region, and one endpoint of the
confidence interval is exactly equal to one of the equivalence bounds
(see situation B in Figure \ref{fig:TOSTSGPV3}) the TOST \emph{p}-value
(which relies on a one-sided test) is always 0.025. The third point
where the SGPV and the \emph{p}-value from the TOST procedure should
overlap is where the 95\% CI falls completely outside of the equivalence
range, but one endpoint of the confidence interval is equal to the
equivalence bound (see situation C in Figure \ref{fig:TOSTSGPV3}), when
the \emph{p}-value will always be 0.975. Note that this situation is in
essence a minimum-effect test (Murphy, Myors, \& Wolach, 2014). The goal
of a minimum-effect is not just to reject a difference of zero, but to
reject the smallest effect size of interest (i.e., the equivalence
bounds). The SGPV summarizes the information from an equivalence test
and a minimum-effect test. These can be two relevant questions to ask,
although it often makes sense to combine an equivalence test and a
null-hypothesis test instead (Lakens et al., 2018).

@@

We can examine this situation by calculating the SGPV and performing the
TOST for a situation where sample sizes are small and the equivalence
range is narrow, such that the CI is more than twice as large as the
equivalence range (see Figure \ref{fig:TOSTSGPV6}). When the two
statistics are plotted against each other we can see where they are
unrelated (indicated by straight lines in the curve, see Figure
\ref{fig:TOSTSGPV7}). We see the SGPV is 0.5 for a range of observed
means where the \emph{p}-value from the equivalence test still varies.
It should be noted that in these calculations the \emph{p}-values for
the TOST procedure are \emph{never} smaller than 0.05 (i.e., they do not
get below 0.05 on the y-axis). In other words, we cannot conclude
equivalence based on any of the observed means. This happens because the
TOST \emph{p}-value is smaller than 0.05 only when the 90\% CI falls
completely between the upper and lower equivalence bounds. However, we
are examining a scenario where the 90\% CI is so wide that it never
falls completely within the two equivalence bounds, and thus the
equivalence test is never significant. As Lakens (2017) notes:
\enquote{in small samples (where CIs are wide), a study might have no
statistical power (i.e., the CI will always be so wide that it is
necessarily wider than the equivalence bounds).} None of the
\emph{p}-values based on the TOST procedure are below 0.05, and thus, in
the long run we have 0\% power.

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/TOSTSGPV9-1.pdf}
\caption{\label{fig:TOSTSGPV9}Comparison of \emph{p}-values from TOST (black
line) and SGPV (dotted grey line) across a range of observed sample
means (x-axis). The sample size is small (n = 10), but because the sd is
half as big as in Figure \ref{fig:TOSTSGPV7} (1 instead of 2) the CI is
less than twice as wide as the equivalence range (set to -0.4 to 0.4).
The SGPV is not set to 0.5 (horizontal light grey line) but reaches a
maximum slightly above 0.5 across a range of observed means.}
\end{figure}

If we once more plot the two statistics against each other to see where
they are unrelated (indicated by straight lines in the curve), we see
the SGPV is 0.56 for a range of observed means where the \emph{p}-value
from the equivalence test still varies (Figure \ref{fig:TOSTSGPV10}).

@@

To conclude this section, there are three situations where the
\emph{p}-value from the TOST procedure is unrelated to the SGPV. In all
these situations the \emph{p}-value for the equivalence test
differentiates tests with different means, but the SGPV does not.
Therefore, as a purely continuous descriptive statistic, the SGPV is
more limited than the \emph{p}-value from the TOST procedure.

@@

This example illustrates the difference between a proportion and a
probability. There is always a 50\% probability of observing a
correlation smaller or larger than the true correlation, but the SGPV
for this situation depends on how far away the observed correlation is
from 0. The further away from 0, the larger the SGPV when the observed
mean falls on the equivalence bound. The SGPV is the proportion of
values in a 95\% confidence interval that overlap with the equivalence
range, but not the probability that these values will be observed. In
the most extreme case (i.e., a sample size of 4, and equivalence bounds
set to r = -0.99 and 0.99, with an observed correlation of 0.99) 58.10\%
of the confidence interval overlaps with the equivalence range, even
though in the long run only 50\% of the correlations observed in the
future will fall in this range. It should be noted that in larger sample
sizes the SGPV is closer to 0.5 whenever the observed correlation falls
on the equivalence bound, but this extreme example nevertheless clearly
illustrates the difference between question the SGPV answers, and the
question a \emph{p}-value answers. The conclusion of this section on
asymmetric confidence intervals is that a SGPV of 1 or 0 can still be
interpreted as a \emph{p} \textless{} 0.025 or \emph{p} \textgreater{}
0.975 in an equivalence test, since the SGPV and \emph{p}-value for the
TOST procedure are always directly related at these values. Although
Blume et al. (2018) state that \enquote{the degree of overlap conveys
how compatible the data are with the null premise} this definition of
what the SGPV provides does not hold for asymmetric confidence
intervals. Although a SGPV of 1 or 0 can be directly interpreted, a SGPV
between 0 and 1 is not interpretable as \enquote{compatibility with the
null hypothesis}. Indeed, Blume and colleagues write in the supplemental
material that \enquote{The magnitude of an inconclusive
second-generation \emph{p}-value can vary slightly when the effect size
scale is transformed. However definitive findings, i.e.~a \emph{p}-value
of 0 or 1 are \emph{not} affected by the scale changes.}

@@

When introducing a new statistical method, it is important to compare it
to existing approaches and specify its relative strengths and
weaknesses. First of all, the SGPV is a descriptive statistic (unlike
the \emph{p}-value that is calculated for an equivalence test, which is
an inferential statistic) that provides the proportion of overlap of the
confidence interval and the equivalence range. Even though a SGPV of 1
or 0 has a clear interpretation (we can reject effects outside or inside
the equivalence range), intermediate values are not as easy to interpret
(especially for effects that have asymmetric confidence intervals). In
one sense, they are what they are (the proportion of overlap), but it
can be unclear what this number tells us about the data we have
collected. This is not too problematic, since the main use of the SGPV
(e.g., in all examples provided by Blume and colleagues) seems to be to
examine whether the SGPV is 0, 1, or inconclusive. This interpretation
of a SGPV as allowing researchers to reject the null, reject the
presence of a meaningful effect, or remaining inconclusive is very
similar to the Neyman-Pearson interpretation of combining a
null-hypothesis test and an equivalence test (Lakens et al., 2018). The
difference is that where a SGPV of 1 can be interpreted as \emph{p}
\textless{} .025, equivalence tests provide exact \emph{p}-values, and
they continue to differentiate between for example \emph{p} = 0.048 and
\emph{p} = 0.002. Whether this is desirable depends on the perspective
that is used. From a Neyman-Pearson perspective on statistical
inferences the main conclusion is based on whether or not
\(p < \alpha\), and thus an equivalence test and SGPV can be performed
by simply checking whether the confidence interval falls within the
equivalence range, just as a null-hypothesis test can be performed by
checking whether the confidence interval contains zero or not. At the
same time, it is recommended to report exact \emph{p}-values (American
Psychological Association, 2010), and exact \emph{p}-values might
provide information of interest to readers about how precisely how
surprising the data is under the null model. Equivalence tests combined
with null-hypothesis significance tests also allow researchers to
conclude an effect is significant \emph{and} equivalent (i.e.,
statistically different from zero, but also too small to be considered
meaningful). Thus, the SGPV is used to classify results into one of
three possible categories (with the data falling inside or outside the
equivalence range, or being inconclusive), while equivalence tests
combined with null-hypothesis tests classify results into four possible
categories.

@@

As a more extreme example of the peculiar behavior of the \enquote{small
sample correction} as currently implemented in the calculation of the
SGPV see Figure \ref{fig:TOSTSGPV13} below. In this figure observed
correlations (from a sample size of 10) from -1 to 1 are tested against
an equivalence range from r = 0.4 to r = 0.8. We can see the SGPV has a
peculiar shape because it is set to 0.5 for certain observed
correlations, even though there is no risk of a \enquote{misleading}
SGPV in this range. This example suggests that the current
implementation of the \enquote{small sample correction} could be
improved. If, on the other hand, the SGPV is mainly meant to be
interpreted when it is 0 or 1, it might be preferable to simply never
apply the \enquote{small sample correction}.

Blume et al. (2018) argue that the SGPV has improved error control, in
that the conclusion that the 95\% confidence interval lies completely
outside the equivalence range around zero will necessarily occur less
frequently than a Type I error in a null-hypothesis test (where the 95\%
confidence interval only needs to not overlap with zero). However, the
SGPV has a \emph{lower} error rate than a null-hypothesis test, not a
\emph{more accurate} error rate. In a Neyman-Pearson perspective (which
forms the basis of equivalence tests) the goal is not to end up with an
error rate that is as low as possible (to achieve that, one simply
adjusts the alpha level), but with a decision procedure that, when
applied, yields a desired error rate with high accuracy. We show here
that the SGPV and equivalence tests are directly related, and so are
their error rates. The TOST procedure uses a 90\% confidence interval
because it is based on two \emph{one-sided} tests, and thus the
confidence interval is 1-2\(\times\)\(\alpha\), and because both tests
need to be significant to conclude equivalence, no adjustment for
multiple comparisons is needed. The TOST procedure always has higher
power to declare equivalence than the SGPV, which relies on a 95\%
confidence interval (but note that there is no reason to always set the
alpha level to 0.05).
